---
title: "GPoM : PreProcessing"
author: "Sylvain Mangiarotti & Mireille Huc"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig.caption: yes
  pdf_document:
    number_sections: no
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
load package: "`r library(GPoM)`"
---

## Pre-processing for global modelling

To apply the global modelling approach,
a careful preprocessing may be necessary
because time series derivatives are required
during the modelling process.
Derivatives will be computed automatically
when using the `gPoMo` function.
To avoid problems, a careful preprocessing
should be applied beforehand to have more chance
to obtain global models.
This vignette aims to show simple examples
of time series preprocessing for global modelling
application.
In particular, we show the usefulness to resample
the data (in order to improve the quality of the
derivatives) and some of the limitations that may
be found, especially when considering derivatives
of high degree.

In practice, other types of preprocessing may be necessary
depending on the quality of the data set (time sampling
and time series length length, level of noise, presence
of gaps, etc.).
The required preprocessing may also depend on other factors
such as: the dimension of underlying dynamics, the number of
observed time series, the degree of observability, etc.


## Single time series

When only one single time series is observed,
a global model of canonical form may be expected:

$dX_1/dt = X_2$

$dX_2/dt = X_3$

$...$

$dX_n/dt = F(X_1, X_2, ..., X_n)$.

with $X_1$ the observed variable,
$X_2$, $X_3$, etc. its successive derivatives.
The GPoM package is based on polynomial formulations.
For the present formulation, its aim is to obtain a
polynomial approximation of the function $F$.
An approximation of $F$ can be obtained using the `gPoMo`
function (see vignette `III_Modelling`).
In practice, the canonical form will require the
computations of $n$ derivatives to have the following
$n + 1$ time series:
$(X_1, X_2, ..., X_n, dX_n/dt)$.

Chaotic systems are well adapted to illustrate the skills
of the global modelling approach because they are not trivial
study cases: they are nonlinear and their behavior is unpredictable
at long term.
The Rössler system introduced by Otto Rössler in 1976
is used here^[O. Rössler, 1976. An Equation for Continuous Chaos,
Physics Letters, 57A(5), p. 397-398.].
This system is defined as:

$dx/dt = - y - z$

$dy/dt = x + a y$

$dz/dt = b + z (x - c)$.

One of the interests of this system is to have
different degrees of observability depending
on the variable considered.
This degree of observability results from the system
nonlinearities.
The question of observability is of first importance
for global modelling since, depending on it,
the model reconstruction may be easy, hard
or completely impossible^[C. Letellier, L. A. Aguirre,
& J. Maquet, 2005. Relation between observability and
differential embeddings for nonlinear dynamics,
Phys. Rev. E 71(6), 066213.].
Time series of such equations can be easily generated
with the GPoM package (see vignette `I_Generate`).
A ready to use data set of the Rössler-1976 system is
also provided in the package that can be directly loaded:

```{r, eval = TRUE, fig.align='center'}
    # load data
    data("Ross76")
    # plot
    tin <- Ross76[,1]
    data <- Ross76[,2:4]
    plot(tin, data[,1],
         xlab = 't', ylab = 'x(t)', main = 'Original time series',
         type='l', col = 'lightgray')
```

The time series $x(t)$ here plotted shows an irregular
behavior and will be used hereafter to illustrate
the global modelling in a canonical form.
To illustrate the problem of time sampling, the time
series is undersampled as follows:

```{r, eval = TRUE, fig.align='center'}
    plot(tin, data[,1],
         xlab = 't', ylab = 'x(t)', main = 'Undersampled time series',
         type='l', col = 'lightgray')
    # subsampling:
    Ross76us <- Ross76[seq(1,4000,by=50),]
    # plot
    tus <- Ross76us[,1]
    datus <- Ross76us[,2:4]
    lines(tus, datus[,1], type='p', col='red', cex = 0.6)
    lines(tus, datus[,1], type='l', col='red')
```

The subsampling leads to loose some information.
However, the general shape is obviously kept.

A spline can be used to resample the signal at the original
sampling rate.

```{r, eval = TRUE, fig.align='center'}
    plot(tin, data[,1],
         xlab = 't', ylab = 'x(t)', main = 'Resampled time series',
         type='l', col = 'lightgray')
    # subsampling:
    Ross76us <- Ross76[seq(1,4000,by=50),]
    # plot
    tus <- Ross76us[,1]
    datus <- Ross76us[,2:4]
    lines(tus, datus[,1], type='p', col='red', cex = 0.6)
    lines(tus, datus[,1], type='l', col='red')
    #
    xout <- seq(min(tus), max(tus), by = 0.01)
    rspl <- spline(tus, y = datus[,1], method = "fmm", xout = xout)
    # plot resampled
    lines(rspl$x, rspl$y, type='l', col='green')
```

In the present case, the spline resampling appears
quite efficient to retrieve the original signal.
Note that alternative methods may be preferred depending on 
the dynamical behavior considered, on the rate of subsampling,
on the signal quality, etc.)

The `drvSucc` function of the GPoM package aims to
estimate the derivatives. It is based on the algorithm
introduced by Savitzky and Golay in 1964^[A. Savitzky & M. J. Golay, 1964.
Smoothing and differentiation of data by simplified least squares
procedures, Anal. Chem. 36(8), 1627–1639.].
This function is applied to the three time series
presented upper:
(1) the original Rössler time series,
(2) its undersampling,
and (3) its resampling to the original sampling using splines.

```{r, eval = TRUE, fig.align='center'}
    # from original signal
    drv1 <- drvSucc(tin, data[,1], nDeriv=2)
    # from undersampled signal
    drv2 <- drvSucc(tus, datus[,1], nDeriv=2)
    # from resampled signal
    drv3 <- drvSucc(rspl$x, rspl$y, nDeriv=2)
```

The obtained time series for the first and second derivatives
are plotted hereafter.
Obviously, much of the original signal (in black) is lost when
computing the derivatives from the undersampled time series
(in red).
Contrarily, much of the original signal is retrieved when
computed from the resampled signal (in green), which clearly
illustate the usefulness to resample the signal with
appropriate algorithms before the computation of the
derivatives.

```{r, eval = TRUE, fig.show='hold'}
    # plot resulting output as a function of time
    # output variable (smoothed) and its derivatives
#    plot(drv1$tout, drv1$seriesDeriv[,1], type='l', xlab = 'time', ylab = 'x(t)')
#    lines(drv2$tout, drv2$seriesDeriv[,1], type='p', cex = 0.6, col = 'red')
#    lines(drv3$tout, drv3$seriesDeriv[,1], type='l', col = 'green')
    # first derivative
    plot(drv1$tout, drv1$seriesDeriv[,2], type='l', xlab = 'time', ylab = 'dx(t)/dt')
    lines(drv2$tout, drv2$seriesDeriv[,2], type='p', cex = 0.6, col = 'red')
    lines(drv3$tout, drv3$seriesDeriv[,2], type='l', col = 'green')
    # second derivative
    plot(drv1$tout, drv1$seriesDeriv[,3], type='l', xlab = 'time', ylab = 'd²x(t)/dt²')
    lines(drv2$tout, drv2$seriesDeriv[,3], type='p', cex = 0.6, col = 'red')
    lines(drv3$tout, drv3$seriesDeriv[,3], type='l', col = 'green')
```

Phase portraits can also be compared.
Two projections are shown here after:
$(x, dx/dt)$ and $(x, d²x/dt²)$.
Obviously, quite much of the information of the original
phase portraits (plotted in black) is lost when the phase
portrait is reconstructed from the undersampled time series
(in red).
Contrarily, the projection $(x, dx/dt)$ of the phase portrait
is retrieved almost perfectly when reconstructed from the
resampled time series (in green). The second phase portrait
$(x, d²x/dt²)$ shows that some drawbacks were introduced
by the resampling, although this processing clearly
improved the signal compared to the undersampled
reconstruction.

```{r, eval = TRUE, fig.show='hold'}
    # output variable (smoothed) and its derivatives
     plot(drv1$seriesDeriv[,1], drv1$seriesDeriv[,2], type='l', xlab = 'x(t)', ylab = 'dx(t)/dt')
    lines(drv2$seriesDeriv[,1], drv2$seriesDeriv[,2], type='l', col = 'red')
    lines(drv3$seriesDeriv[,1], drv3$seriesDeriv[,2], type='l', col = 'green')
    # first derivative
#     plot(drv1$seriesDeriv[,1], drv1$seriesDeriv[,3], type='l', xlab = 'x(t)', ylab = 'd²x(t)/dt²')
#    lines(drv2$seriesDeriv[,1], drv2$seriesDeriv[,3], type='l', col = 'red')
#    lines(drv3$seriesDeriv[,1], drv3$seriesDeriv[,3], type='l', col = 'green')
    # second derivative
     plot(drv1$seriesDeriv[,2], drv1$seriesDeriv[,3], type='l', xlab = 'dx(t)/dt', ylab = 'd²x(t)/dt²')
    lines(drv2$seriesDeriv[,2], drv2$seriesDeriv[,3], type='l', col = 'red')
    lines(drv3$seriesDeriv[,2], drv3$seriesDeriv[,3], type='l', col = 'green')
```

This preprocessing illustrates (1) the usefulness to have
a proper sampling of the observed process; (2) the usefulness
to resample the observed time series at a better time sampling
to improve the estimates of the derivatives;
(3) some of the limitations reached for the derivatives of
higher degree.
Subsampling and resampling may have a direct impact when
applying the global modelling techique and should thus be
considered carefully when using the approach.
Since the derivatives will be computed automatically when
using the `gPoMo` function, it is important to keep in mind
that their quality will directly depend on the characteristic
of the time series provided as an input.


## Multiple time series

When several time series are available,
global models of the following form can be expected:

$dX_1/dt = f_1(X_1, X_2, ..., X_n)$

$dX_2/dt = f_2(X_1, X_2, ..., X_n)$

$...$

$dX_n/dt = f_n(X_1, X_2, ..., X_n)$.

where the $X_i$, are the observed variables.
The GPoM package can be used to obtain a polynomial
approximation of the functions $f_i$ (see vignette `III_Modelling`).
This formulation will require the computations of the
first derivatives of each time series in order to have
$(X_1, X_2, ..., X_n, dX_1/dt, dX_2/dt, ... dX_n/dt)$.

Here again the Rössler system (1976) is used to
illustrate the skills of the global modelling approach.

```{r, eval = TRUE, fig.align='center'}
    # load
    data("Ross76")
    # plot
    tin <- Ross76[,1]
    data <- Ross76[,2:4]
    plot(tin, data[,1],
         ylim = c(-6.5,12),
         type='l', col='blue', xlab = 'time', ylab = 'x(t), y(t), z(t)')
    lines(tin, data[,2], type='l',  col='orange')
    lines(tin, data[,3], type='l',  col='brown')
```

The phase space of the original Rössler system can be
reconstructed for the following set of time series
$x(t)$, $y(t)$ and $z(t)$:
The following reconstructions correspond to projections $(x,y)$ and $(x,z)$:

```{r, eval = TRUE, fig.show='hold'}
    plot(data[,1], data[,2], type='l', xlab = 'x(t)', ylab = 'y(t)')
    #plot(data[,1], data[,3], type='l', xlab = 'y(t)', ylab = 'z(t)')
    plot(data[,2], data[,3], type='l', xlab = 'x(t)', ylab = 'z(t)')
```

For practical reasons, observations are often undersampled:

```{r, eval = TRUE, fig.align='center'}
    # plot
    tin <- Ross76[,1]
    data <- Ross76[,2:4]
    plot(tin, data[,1],
         ylim = c(-6.5,12),
         type='l', col='lightgray', xlab = 'time', ylab = 'x(t)')
    lines(tin, data[,2], type='l',  col='lightgray', xlab = 'time', ylab = 'y(t)')
    lines(tin, data[,3], type='l',  col='lightgray', xlab = 'time', ylab = 'z(t)')

    # undersampled data
    #
    # subsampling:
    Ross76us <- Ross76[seq(1,4000,by=75),]
    # plot
    tus <- Ross76us[,1]
    datus <- Ross76us[,2:4]
    lines(tus, datus[,1], type='p', cex = 0.5, col='blue')
    lines(tus, datus[,2], type='p', cex = 0.5, col='orange')
    lines(tus, datus[,3], type='p', cex = 0.5, col='brown')
```

Subsampling may lead to loose an important part of 
the original signal.
This lost may vary from one observable to another
(for example, several picks of the time series are
missrepresented by the undersampled signal), and
this effect will become stronger when
considering derivatives of higher degrees (see the
first Section of the present vignette).
It can be useful in such case to resample the time
series at a higher time resolution.
Resampling can be performed with splines as 
exemplified upper for single time series.
Alternative methods may be preferred depending on 
the complexity of the behavior under study, on the 
degree of subsampling, on the level of noise, etc.
Here, this resampling appears efficient enough to obtain
a good reconstruction of the phase space as illustrated
by the following projections $(x,y)$ and $(x,z)$.

```{r, eval = TRUE, fig.show='hold'}
    xout <- seq(min(tus), max(tus), by = 0.01)
    rsplx <- spline(tus, y = datus[,1], method = "fmm", xout = xout)
    rsply <- spline(tus, y = datus[,2], method = "fmm", xout = xout)
    rsplz <- spline(tus, y = datus[,3], method = "fmm", xout = xout)
    # plot resampled
    plot(data[,1], data[,2], type='l', xlab = 'x(t)', ylab = 'y(t)')
    lines(rsplx$y, rsply$y, type='l', col='green')
#    plot(data[,1], data[,3], type='l', xlab = 'y(t)', ylab = 'z(t)')
#    lines(rsplx$y, rsplz$y, type='l', col='green')
    plot(data[,2], data[,3], type='l', xlab = 'x(t)', ylab = 'z(t)')
    lines(rsply$y, rsplz$y, type='l', col='green')
```

Discrepancies related to the derivatives
can be easily observed considering the derivative
phase portraits.
The skills of resampling is obviously good for variable $y$:

```{r, eval = TRUE, fig.align='center'}
    # derivatives
    # from original signal
    drv1x <- drvSucc(tin, data[,1], nDeriv=2)
    drv1y <- drvSucc(tin, data[,2], nDeriv=2)
    drv1z <- drvSucc(tin, data[,3], nDeriv=2)
    # from resampled signal
    drv3x <- drvSucc(rsplx$x, rsplx$y, nDeriv=2)
    drv3y <- drvSucc(rsply$x, rsply$y, nDeriv=2)
    drv3z <- drvSucc(rsplz$x, rsplz$y, nDeriv=2)
    # phase portraits plot for y:
    # (y, dy/dt) projection of the original signal (in black) and resampled signal (in green)
    plot(drv1y$seriesDeriv[,1], drv1y$seriesDeriv[,2], type='l', xlab = 'y', ylab = 'dy/dt')
    lines(drv3y$seriesDeriv[,1], drv3y$seriesDeriv[,2], type='l', col='green')
```

Resampling appears less efficient for variables $x$ and $z$:

```{r, eval = TRUE, fig.show='hold'}
    # phase portraits plot for x and z:
    # (x, dx/dt) projection of the original signal (in black) and resampled signal (in green)
    plot(drv1x$seriesDeriv[,1], drv1x$seriesDeriv[,2], type='l', xlab = 'x', ylab = 'dx/dt')
    lines(drv3x$seriesDeriv[,1], drv3x$seriesDeriv[,2], type='l', col='green')
    # phase portraits plots
    # (z, dz/dt) projection of the original signal (in black) and resampled signal (in green)
    plot(drv1z$seriesDeriv[,1], drv1z$seriesDeriv[,2], type='l', xlab = 'z', ylab = 'dz/dt')
    lines(drv3z$seriesDeriv[,1], drv3z$seriesDeriv[,2], type='l', col='green')
```

Although resampling with splines may not permit to reconstruct
the original signal perfectly, from a practical point of view,
it appears necessary in many cases, and it is in practice a rather
efficient way to deal with subsampling.
This type of resampling could be efficiently used for
applying the global modelling technique both from single time
series^[S. Mangiarotti, L. Drapeau, & C. Letellier, 2014.
Two chaotic global models for cereal crops cycles observed from
satellite in Northern Morocco, Chaos, 24, 023130.],
and multiple time series^[S. Mangiarotti, 2015. Low dimensional
chaotic models for the plague epidemic in Bombay (1896–1911),
Chaos, Solitons, Fractals 81(A), 184–196.] ^[S. Mangiarotti,
M. Peyre & M. Huc, 2016. A chaotic model for the epidemic
of Ebola virus disease in West Africa (2013–2016).
Chaos, 26, 113112.].

## Next step

Once the time series have been carefuly prepared, the global
modelling technique can be tried. This is presented in
the next vignette `III_Modelling`.
