---
title: "GPoM : Visualization of the outputs"
author: "Sylvain Mangiarotti & Mireille Huc"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig.caption: yes
  pdf_document:
    number_sections: no
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
load package: "`r library(GPoM)`"
---

## Model vizualization

The GPoM package permits to explore large numbers
of models in polynomial ODEs in order to identify
what model is able to reproduce an observed
behaviour.
The present vignette aims to explain how to retrieve
the information obtained from the `gPoMo` function.

In practice, the `gPoMo` function will generate
a list of variables hereafter called `outputGPoM`:

```{r, eval = TRUE}
  data("Ross76")
  # time vector
  tin <- Ross76[seq(1, 3000, by = 8), 1]
  # single time series
  data <- Ross76[seq(1, 3000, by = 8), 3]
  # global modelling
  # results are put in list outputGPoM
  outputGPoM <- gPoMo(data, tin=tin, dMax = 2, nS=c(3),
                      show = 0, method = 'rk4',
                      nPmax = 12, IstepMin = 400, IstepMax = 401)
```

To test the potential of a given model to reproduce
the observed dynamics, the numerical integrability
of each model is tested and the resulting behaviour
may either diverge, converge to a trivial attractor
(fixed points, periodic cycles) or to a chaotic attractor.

The aim of the present vignette is show how to have
a rapid overview of the models explored or obtained
with the `gPoMo` function.

In order to keep a memory of the analyzed signal, the data
used as an input are kept in `$tin` (for the input time)
and `$inputdata` (for the input time series):

```{r, eval = TRUE, fig.show='center'}
  # plot data and models time series
  plot(outputGPoM$tin, outputGPoM$inputdata,
       xlab='t', ylab='y(t)', cex = 0.4, main = 'Original time series')
```

The filtered version obtained when computing
the derivatives is in `$tfilt` (the vector time of
the filtered time series which length is lower due
to some lost at the boudaries) and `$filtdata`,
a matrix with the filtered time series (first column)
and its successive derivatives (second and next columns).

```{r, eval = TRUE, fig.show='center'}
  # plot data and models time series
  plot(outputGPoM$tin, outputGPoM$inputdata,
       xlab='t', ylab='y(t)', cex = 0.4, main = 'with filtered time series')
  lines(outputGPoM$tfiltdata, outputGPoM$filtdata[,1], type='l', col='green')
```

The original differential phase portrait can thus
be plotted as follows:

```{r, eval = TRUE, fig.show='hold'}
  # plot data and models time series
  plot(outputGPoM$filtdata[,1], outputGPoM$filtdata[,2],
       xlab='y', ylab='dy/dt', main = 'First projection',
       type = 'l', cex = 0.4)
  plot(outputGPoM$filtdata[,1], outputGPoM$filtdata[,3],
       xlab='y', ylab='d²y/dt²', main = 'Second projection',
       type = 'l', cex = 0.4)
```

The information about the tested models are in the other
variables: `$okMod`, `$stockoutreg` and `$models`.

Variable `$okMod` provides information about the obtained
numerical ingration of all the tested models: values are
chosen equal to zero when the integrated trajectory
diverged during the numerical integration,
equal to 1 when the model could not be categorized
(meaning that the model is still under categorizing
test).

Since we are mainly interested in non-diverging models,
these should have been identified in the `$okMod` parameter
such as: `$okMod != 0`. Note that diverging models may be
identified as non-diverging if the numerical integration
was not performed on a sufficiently long duration.
The number of `$okMod` identified as non-diverging
(based on the chosen maximum integration steps `IStepMax`)
is obtained by:

```{r, eval = TRUE}
  ######################
  # manual    ##########
  ######################
  # How many models could not be identified as non-diverging
  sum(outputGPoM$okMod)
```

The reference number of the uncategorized models can be
retrieved as follows:

```{r, eval = TRUE}
  # what is the reference number of these models?
  which(outputGPoM$okMod == 1)
```

The numerical integration of the tested models is kept
in variable `$stockoutreg`. For example, it can be plotted
for models #7 and #9:

```{r, eval = TRUE, fig.show='hold'}
  # plot data and models time series
  plot(outputGPoM$tin, outputGPoM$inputdata,
       xlab='t', ylab='y(t)', cex = 0.7, main = 'Model 7')
  lines(outputGPoM$tfiltdata, outputGPoM$filtdata[,1], type='l', col='green')
  lines(outputGPoM$tout, outputGPoM$stockoutreg$model7[,1], type='l', col='red')
  #
  plot(outputGPoM$tin, outputGPoM$inputdata,
       xlab='t', ylab='y(t)', cex = 0.7, main = 'Model 9')
  lines(outputGPoM$tfiltdata, outputGPoM$filtdata[,1], type='l', col='green')
  lines(outputGPoM$tout, outputGPoM$stockoutreg$model9[,1], type='l', col='red')
```

Note that only one iteration of the model test
was performed (since `IstepMin` $* 2 <$ `IstepMax`).
As a consequence, the same initial conditions
are used for both filtered data and model reconstruction:

```{r, eval = TRUE}
  head(outputGPoM$filtdata[1:3], 1)
  head(outputGPoM$stockoutreg$model7, 1)
```

(note that `$filtdata` has one more derivative
which was necessary to identify the model
parameters).

Original and models phase portraits can also
be plotted for comparison as follows:

```{r, eval = TRUE, fig.show='hold'}
  # plot data and models phase portraits
  plot(outputGPoM$filtdata[,1], outputGPoM$filtdata[,2], type='l', col='green',
       xlab='y(t)', ylab='dy(t)/dt', main = 'Model 7')
  lines(outputGPoM$stockoutreg$model7[,1], outputGPoM$stockoutreg$model7[,2],
        type='l', col='red')
  #
  plot(outputGPoM$filtdata[,1], outputGPoM$filtdata[,2], type='l', col='green',
       xlab='y(t)', ylab='dy(t)/dt', main = 'Model 9')
  lines(outputGPoM$stockoutreg$model9[,1], outputGPoM$stockoutreg$model9[,2],
        type='l', col='red')
```

Finally, models coefficients are all kept in the sublist
`$models`. These are separated in `$models$mToTest#`
where `#` should be replaced by the model number
(the list of the models that were tested with
the `gPoMo` algorithm) and `$models$model#`
(the list of uncategorized models).
In both cases, models coefficients are in a matrix
of dimension `pMax * nVar` with `pMax` the maximal
number of parameter (see vignette `I_Generate`) and
`nVar` the number of variables (also called model
dimension).
The corresponding equations can be edited directly
using the `visuEq` function as:

```{r, eval = TRUE}
  nVar <- dim(outputGPoM$models$model7)[2]
  pMax <- dim(outputGPoM$models$model7)[1]
  dMax <- p2dMax(nVar,pMaxKnown = pMax)
  # See equations
  visuEq(nVar, dMax, outputGPoM$models$model7, approx = 2)
  #
  visuEq(nVar, dMax, outputGPoM$models$model9, approx = 2)
```

By default, a large number of digits is provided for each
parameter.
To make its reading easier, this number can be controlled
by using the optinonal parameter `approx = TRUE`, 
the minimum number of digit will be kept (the same for
all the parameters) but keeping all the terms in the
polynomial.
The number of digits to add to the minimum number of digits
required can also be chosen manually.

The number of terms of each model can be obtained as follows:

```{r, eval = TRUE}
  # for model #7
  sum(outputGPoM$models$model7 != 0)
  # for model #9
  sum(outputGPoM$models$model9 != 0)
```

Since these models have a canonical structure, it may also
be interesting to know the number of terms in the last
equation (which polynomial defines the model), such as:

```{r, eval = TRUE}
  # for model #7
  colSums(outputGPoM$models$model7 != 0)
  # for model #9
  colSums(outputGPoM$models$model9 != 0)
```

Finally, to have an overview of the overall outputs
of `gPoMo` at a glance for all the models, the following
command can be used:

```{r, eval = FALSE}
  ######################
  # automatic ##########
  ######################
  visuOutGP(outputGPoM)
```

## Next step

An overview of all the main elements of the global
modelling technique and GPoM tools were presented
in the previous vignettes.
An illustration of model validation will now be
presented in the last vignette `V_Predictability`.
