---
title: "GPoM : 1 Conventions"
author: "Sylvain Mangiarotti & Mireille Huc"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: no
vignette: >
  %\VignetteIndexEntry{GPoM : 1 Conventions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
load package: "`r library(GPoM)`"
---

The Generalized Global Polynomial Modelling (GPoM)
package allows a generic
formulation of any Ordinary Differential Equations (ODEs)
in polynomial form. The aim of the present vignette
`1 Conventions` is to introduce briefly the way to describe
a set of polynomial ODEs with GPoM and to show how
to perform its numerical integration^[Mangiarotti S,
Le Jean F, Chassan M, Drapeau L, Huc M. 2018.
GPoM: Generalized Polynomial Modelling. Version 1.1.
*Comprehensive R Archive Network*.
<https://cran.r-project.org/web/packages/GPoM>.].


## Conventions used to describe a polynomial

The polynomial description is based on a convention defined
by the function `regOrd` that provides the order of the polynomial
terms.
This convention depends on the model dimension (that is the number
`nVar` of state variables), and on the maximum polynomial degree
used for the formulation (defined by the parameter `dMax`).
This order can be visualized using the `poLabs` function.
For instance, for `nVar = 3` and `dMax = 2`, the convention
used to formulate a polynomial will be:


```{r}
nVar = 3
dMax = 2
poLabs(nVar = nVar, dMax = dMax)
```

This formulation has `pMax = 10` terms:

```{r, echo=FALSE}
nVar = 3
dMax = 2
```
```{r}
pMax =  d2pMax(nVar, dMax)
```
Based on this convention, one single ordinary differential equation (ODE) in polynomial
form with `nVar` variables and of maximum polynomial degree `dMax` can be formulated
as one single vector using the convention given by `regOrd(nVar, dMax)`.
As an example, the equation :

$dX_1/dt = 1 + 2 X_1 - 3 X_1X_3 + 4 X_2^2$

has the three variables $X_1$, $X_2$ and $X_3$
(it thus requires at least `nVar = 3`) and is of maximum polynomial
degree two due to terms $X_1X_3$ and $X_2^2$
(it thus requires at least `dMax = 2`).
Following the convention defined by `poLabs(nVar = 3, dMax = 2)`,
it will require the definition of the following vector of parameters:

```{r}
param <- c(1, 0, 0, 0, 0, 4, 2, -3, 0, 0)
```

Indeed:
```{r}
nVar = 3
dMax = 2
cbind(param, poLabs(nVar, dMax))
```
The same convention will be used for any other equation.
Note that, by default, the notation used for the variables
is `X1`, `X2`, etc.
However, to facilitate the analysis, alternative notations may
also be used using the optional parameter `Xnote`:

```{r}
poLabs(3, 2, Xnote = 'y')
```

or for a full choice of the notation:

```{r}
poLabs(3, 2, Xnote = c('x','W','y'))
```


## Definition of a set of polynomial ODE

A set of $N$ equations will require the definition 
of $N$ parameter vectors and will thus be represented
by a matrix of `pMax` lines by `nVar` columns.
For example, the Rössler system^[O. Rössler, An Equation for Continuous Chaos,
Physics Letters, **57A**(5), 1976, 397-398]
is defined by a set of three equations


$dx/dt = - y - z$

$dy/dt = x + a y$

$dz/dt = b + z (x - c)$.


For $(a = 0.52, b = 2, c = 4)$, this system can be decribed
by three vectors (one for each equation)
```{r}
# parameters
a = 0.52
b = 2
c = 4
# equations
Eq1 <- c(0,-1, 0,-1, 0, 0, 0, 0, 0, 0)
Eq2 <- c(0, 0, 0, a, 0, 0, 1, 0, 0, 0)
Eq3 <- c(b,-c, 0, 0, 0, 0, 0, 1, 0, 0)
```
The model formulation is obtained by concatenating the vectors of
the three equations into one single matrix `K` containing all the
coefficients of the model:

```{r}
K = cbind(Eq1, Eq2, Eq3)
```

The corresponding model equations can be edited in a mathematical
form using the function `visuEq()`

```{r}
visuEq(nVar, dMax, K)
```

By default, the notation used in `visuEq` for the variables
is `X` with an indicative number, such as `X1`, `X2`, etc.
Alternative notation can be used to edit the equations with 
the `visuEq` function using the optional parameter `substit`.
For `substit = 1`, single letters are automatically chosen
such as

```{r}
visuEq(nVar, dMax, K, substit = 1)
```

The notation can be defined also manually, such as:

```{r, eval=TRUE}
visuEq(nVar, dMax, K, substit = c("U", "V", "W"))
```

## Numerical integration

The numerical integration of the model defined by matrix `K`
can be done using the `numicano` function. It requires
the use of the external package `deSolve`.
The following parameters are required as input:

```{r}
# The initial conditions of the system variables
v0 <- c(-0.6, 0.6, 0.4)
# the model formulation K (see former section)
# the number of integration steps `Istep`
nIstep <- 5000
# the time step length `onestep`
onestep = 1/50
# the model dimension `nVar`
nVar = 3
# the maximum polynomial degree `dMax`
dMax = 2
```

The numerical integration is launched as follows:

```{r}
outNumi <- numicano(nVar, dMax, Istep = nIstep, onestep = onestep, KL = K, v0 = v0)
```

The output of the function `numicano` is a list that contains
(1) a memory `$KL` of the model parameters

```{r, eval=FALSE}
outNumi$KL
```

from which `nVar` and `dMax` (required to reformulate the equations)
can be retrieved

```{r, eval = FALSE}
# nVar
dim(outNumi$K)[2]
# dMax
pMax <- dim(outNumi$K)[1]
p2dMax(nVar, pMaxKnown = pMax)
```

and (2) the simulations `$reconstr`.
This matrix has `nVar + 1` columns.
The first one is the time, the other ones correspond
to the variables of the system $(X1, X2, X3, ...)$
(or $(x,y,z)$ to keep the formulation used previously
in the text).

Note that all the other input parameters used in `numicano`
can be retrieved from the outputs:

```{r, eval=FALSE}    
# initial conditions
head(outNumi$reconstr, 1)[2:(nVar+1)]
# time step
diff(outNumi$reconstr[1:2,1])
# number of integration time step
dim(outNumi$reconstr)[1]
```

The simulated time series can be plotted as follows:

```{r, fig.show='hold', fig.align='center'}    
plot(outNumi$reconstr[,1], outNumi$reconstr[,2], type='l',
     main='time series', xlab='t', ylab = 'x(t)')
```

and the plot of the phase portrait as well:


```{r, fig.show='hold', fig.align='center', fig.width=4, fig.height=4}  
plot(outNumi$reconstr[,2], outNumi$reconstr[,3], type='l',
     main='phase portrait', xlab='x(t)', ylab = 'y(t)')
```

## Next steps

The aim of the GPoM package is to retreive ODEs from
time series using global modelling. Such type of modelling
may require a careful data preprocessing.
Simple examples of preprocessing will be given in the next
vignette `2 Preprocessing`.
Examples for applying global modelling to time series
will then be presented in vignette `3 Modelling`.
