---
title: "GPoM : II PreProcessing"
author: "Sylvain Mangiarotti & Mireille Huc"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: no
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{GPoM : II PreProcessing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
load package: "`r library(GPoM)`"
---

## Pre-processing for global modelling

To obtain Ordinary Differential Equations (ODEs)
from time series requires the computation
of its derivative(s).
A careful preprocessing of the observed time series
may thus be necessary before starting the modelling process.
Although derivatives will be computed automatically
when using the `gPoMo` function, some preprocessing
may generally be necessary.
This vignette aims to show simple examples
of time series preprocessing for global modelling
application.
In practice, various types of preprocessing may be
necessary that will depend on the quality of the data set
(time sampling and time series length, level of noise,
presence of gaps, etc.).
In the present vignette, we will mainly focus on
the problem of subsampled time series.
We will illustrate the usefulness to resample
the data (with the aim to improve the quality of the
derivatives) and some of the limitations
that may be found, especially when considering
derivatives of higher degree.

The required preprocessing may also depend on other factors
such as: the dimension of underlying dynamics, the number of
observed time series, the degree of observability, etc.


## Single time series

When only a single time series is observed,
a global model of canonical form may be expected:

$dX_1/dt = X_2$

$dX_2/dt = X_3$

$...$

$dX_n/dt = F(X_1, X_2, ..., X_n)$,

with $X_1$ the observed variable,
and $X_2$, $X_3$, etc. its successive derivatives.
The GPoM package is based on polynomial formulations.
For the canonical formulation, its aim is to obtain a
polynomial approximation of the function $F$.
An approximation of $F$ can be obtained using the `gPoMo`
function (see vignette `III_Modelling`).
In practice, the canonical form will require the
computations of $n$ derivatives in order to have
the following $n + 1$ time series:
$(X_1, X_2, ..., X_n, dX_n/dt)$.

Chaotic systems are well adapted to illustrate the performances
of the global modelling approach because they are not trivial
study cases: they are nonlinear and their behavior is
unpredictable at long term ^[S. Mangiarotti, Modélisation globale
et caractérisation topologique de dynamiques environnementales:
de l'analyse des enveloppes fluides et du couvert de surface
de la Terre à la caractérisation topolodynamique du
chaos, Habilitation to Direct Researches, Université de
Toulouse 3, 2014.].
The Rössler system introduced by Otto Rössler in 1976
is used here^[O. Rössler, 1976. An Equation for Continuous Chaos,
*Physics Letters*, **57A**(5), p. 397-398.] to illustrate
the performances of the tool.
This system is defined as:

$dx/dt = - y - z$

$dy/dt = x + a y$

$dz/dt = b + z (x - c)$.

One of the interests of this system is to have
different degrees of observability depending
on the considered variable.
This degree of observability results from the system
nonlinearities.
The question of observability is of first importance
for global modelling since
the model reconstruction may be either easy, hard
or even completely impossible depending on the degree
of observability^[C. Letellier, L. A. Aguirre,
& J. Maquet, 2005. Relation between observability and
differential embeddings for nonlinear dynamics,
*Physical Review E*, **71**(6), 066213.].
Time series of such equations can be easily generated
with the GPoM package (see vignette `I_Generate`).
A ready-to-use data set of the Rössler-1976 system that
can be directly loaded is also provided in the package:

```{r, eval = TRUE, fig.align='center'}
# load data
data("Ross76")
# plot
tin <- Ross76[,1]
data <- Ross76[,2:4]
plot(tin, data[,1], xlab = 't', ylab = 'x(t)', main = 'Original time series',
     type='l', col = 'gray')
```

The time series $x(t)$ plotted above shows an irregular
behavior and will be used hereafter to illustrate
the global modelling with a canonical formulation.
To illustrate the problem of time sampling, the time
series is subsampled as follows:

```{r, eval = TRUE, fig.align='center'}
plot(tin, data[,1], xlab = 't', ylab = 'x(t)', 
     main = 'Subsampled time series', type='l', col = 'gray')
# subsampling:
Ross76us <- Ross76[seq(1,4000,by=50),]
# plot
tus <- Ross76us[,1]
datus <- Ross76us[,2:4]
lines(tus, datus[,1], type='p', col='red', cex = 0.6)
lines(tus, datus[,1], type='l', col='red')
legend(0,6.5,c("original", "subsampled"), col=c('gray', 'red'),
       lty=c(1,1), pch=c(NA,1), cex=0.6)

```

The subsampling leads to loose some information.
However, the general shape is obviously kept.

A spline can be used to resample the signal at the original
sampling rate.

```{r, eval = TRUE, fig.align='center'}
plot(tin, data[,1], xlab = 't', ylab = 'x(t)', 
     main = 'Resampled time series', type='l', col = 'lightgray')
# subsampling:
Ross76us <- Ross76[seq(1,4000,by=50),]
# plot
tus <- Ross76us[,1]
datus <- Ross76us[,2:4]
lines(tus, datus[,1], type='p', col='red', cex = 0.6)
lines(tus, datus[,1], type='l', col='red')
#
xout <- seq(min(tus), max(tus), by = 0.01)
rspl <- spline(tus, y = datus[,1], method = "fmm", xout = xout)
# plot resampled
lines(rspl$x, rspl$y, type='l', col='green')
legend(0,6.5,c("original", "subsampled", "resampled"), col=c('lightgray', 'red', 'green'), lty=c(1,1,1), pch=c(NA,1, NA), cex=0.6)
```

In the present case, the spline resampling appears
quite efficient to retrieve the original signal.
Note that alternative methods may be preferred (depending on 
the dynamical behavior considered, on the rate of subsampling,
on the signal quality, etc.)

The `drvSucc` function of the GPoM package aims to
estimate temporal derivatives from a time series.
It is based on the algorithm introduced by Savitzky & Golay
in 1964^[A. Savitzky & M. J. Golay, 1964.
Smoothing and differentiation of data by simplified least squares
procedures, *Analytical Chemistry*, **36**(8), 1627-1639.].
This function is here applied to the three time series $x(t)$
presented upper:
(1) the original Rössler time series,
(2) its subsampling,
and (3) its resampling to the original sampling using splines.

```{r, eval = TRUE, fig.align='center'}
# from original signal
drv1 <- drvSucc(tin, data[,1], nDeriv=2)
# from subsampled signal
drv2 <- drvSucc(tus, datus[,1], nDeriv=2)
# from resampled signal
drv3 <- drvSucc(rspl$x, rspl$y, nDeriv=2)
```

The obtained time series of the first and second derivatives
are plotted hereafter.
Obviously, much of the original signal (in black) is lost when
computing the derivatives from the subsampled time series
(in red).
Contrarily, much of the original signal is retrieved when
estimated from the resampled signal (in green), which clearly
illustates the usefulness to resample the signal with
appropriate algorithms before computing the derivatives.

```{r, eval = TRUE, fig.show='hold', fig.align='center'}
# plot resulting output as a function of time
# first derivative
plot(drv1$tout, drv1$seriesDeriv[,2], type='l', xlab = 'time', 
     ylab= expression(dx/dt))
lines(drv2$tout, drv2$seriesDeriv[,2], type='p', cex = 0.6, col = 'red')
lines(drv3$tout, drv3$seriesDeriv[,2], type='l', col = 'green')
legend(0,-5,c("original", "subsampled", "resampled"), 
       col=c('black', 'red', 'green'), lty=c(1,NA,1), pch=c(NA,1, NA), cex=0.6)
# second derivative
plot(drv1$tout, drv1$seriesDeriv[,3], type='l', xlab = 'time', 
     ylab= expression(d^2*x/dt^2))
lines(drv2$tout, drv2$seriesDeriv[,3], type='p', cex = 0.6, col = 'red')
lines(drv3$tout, drv3$seriesDeriv[,3], type='l', col = 'green')
legend(0,-10,c("original", "subsampled", "resampled"), 
       col=c('black', 'red', 'green'), lty=c(1,NA,1), pch=c(NA,1, NA), cex=0.6)
```

Phase portraits can also be compared.
Two projections are shown hereafter:
$(x, dx/dt)$ and $(x, d^2x/dt^2)$.
Obviously, quite much of the information of the original
phase portraits (plotted in black) is lost when the phase
portrait is reconstructed from the subsampled time series
(in red).
Contrarily, the projection $(x, dx/dt)$ of the phase portrait
is retrieved almost perfectly when reconstructed from the
resampled time series (in green). The second phase portrait
$(x, d^2x/dt^2)$ shows that some drawbacks were introduced
by the resampling, although this processing clearly
improved the signal compared to the subsampled
reconstruction.

```{r, eval = TRUE, fig.show='hold', fig.align='center', fig.width=4, fig.height=4}
# output variable (smoothed) and its derivatives
plot(drv1$seriesDeriv[,1], drv1$seriesDeriv[,2], type='l', xlab = 'x(t)', 
     ylab= expression(dx/dt))
lines(drv2$seriesDeriv[,1], drv2$seriesDeriv[,2], type='l', col = 'red')
lines(drv3$seriesDeriv[,1], drv3$seriesDeriv[,2], type='l', col = 'green')
legend(-3,-5,c("original", "subsampled", "resampled"), 
       col=c('black', 'red', 'green'), lty=1, cex=0.5)
# second derivative
plot(drv1$seriesDeriv[,2], drv1$seriesDeriv[,3], type='l', xlab = 'dx(t)/dt', 
     ylab= expression(d^2*x/dt^2))
lines(drv2$seriesDeriv[,2], drv2$seriesDeriv[,3], type='l', col = 'red')
lines(drv3$seriesDeriv[,2], drv3$seriesDeriv[,3], type='l', col = 'green')
legend(1.3,18,c("original", "subsampled", "resampled"), 
       col=c('black', 'red', 'green'), lty=1, cex=0.5)
```

This preprocessing illustrates (1) the usefulness to have
a proper sampling of the observed process; (2) the usefulness
to resample the observed time series at a better time sampling
to improve the estimates of the derivatives;
(3) some of the limitations reached for the derivatives of
higher degree.
Subsampling will have a direct and drastic impact when
applying the global modelling technique.
Resampling can be very efficient to retrieve the original
signal and to obtain a good model^[S. Mangiarotti, 2018.
The global modelling classification technique applied to
the detection of chaotic attractors. *Supplementary Material A* to
"Can the global modelling technique be used for crop classification?"
by S. Mangiarotti, A.K. Sharma, S. Corgne, L. Hubert-Moy, L. Ruiz, M.
Sekhar, Y. Kerr, 2018. *Chaos, Solitons & Fractals*, **106**, 363-378.]

Since the derivatives will be computed automatically when
using the `gPoMo` function, it is important to keep in mind
that their quality will directly depend on the characteristic
of the time series used as input.


## Multiple time series

When several time series are available,
global models of the following form can be expected:

$dX_1/dt = f_1(X_1, X_2, ..., X_n)$

$dX_2/dt = f_2(X_1, X_2, ..., X_n)$

$...$

$dX_n/dt = f_n(X_1, X_2, ..., X_n)$.

where the $X_i$, are the observed variables.
The GPoM package can be used to obtain a polynomial
approximation of the functions $f_i$ (see vignette `III_Modelling`).
This formulation will thus require the computations of the
first derivatives of each time series in order to have
$(X_1, X_2, ..., X_n, dX_1/dt, dX_2/dt, ... dX_n/dt)$.

Here again the Rössler system (1976) is used to
illustrate the performances of the global modelling approach.

```{r, eval = TRUE, fig.align='center'}
# load
data("Ross76")
# plot
tin <- Ross76[,1]
data <- Ross76[,2:4]
plot(tin, data[,1], ylim = c(-6.5,12), xlab = 'time', ylab = '', type='l', col='blue')
lines(tin, data[,2], type='l',  col='orange')
lines(tin, data[,3], type='l',  col='brown')
legend(35,12,c("x(t)", "y(t)", "z(t)"), col=c('blue', 'orange', 'brown'), lty=1, cex=0.8)
```

The phase space of the original Rössler system can be
reconstructed for the following set of time series
$x(t)$, $y(t)$ and $z(t)$:
The following reconstructions correspond to projections $(x,y)$ and $(x,z)$:

```{r, eval = TRUE, fig.show='hold', fig.align='center', fig.width=4, fig.height=4}
plot(data[,1], data[,2], type='l', xlab = 'x(t)', ylab = 'y(t)')
```
```{r, eval = TRUE, fig.show='hold', fig.align='center', fig.width=4, fig.height=4}
plot(data[,2], data[,3], type='l', xlab = 'x(t)', ylab = 'z(t)')
```
For practical reasons (noise reduction, memory size, etc.),
observations may be subsampled:

```{r, eval = TRUE, fig.align='center'}
# plot
tin <- Ross76[,1]
data <- Ross76[,2:4]
plot(tin, data[,1], ylim = c(-6.5,12), type='l', col='lightgray', xlab = 'time', ylab = '')
lines(tin, data[,2], type='l',  col='lightgray')
lines(tin, data[,3], type='l',  col='lightgray')

# subsampled data
#
# subsampling:
Ross76us <- Ross76[seq(1,4000,by=75),]
# plot
tus <- Ross76us[,1]
datus <- Ross76us[,2:4]
lines(tus, datus[,1], type='p', cex = 0.5, col='blue')
lines(tus, datus[,2], type='p', cex = 0.5, col='orange')
lines(tus, datus[,3], type='p', cex = 0.5, col='brown')
legend(35,12,c("x(t)", "y(t)", "z(t)"), col=c('blue', 'orange', 'brown'), pch=1, cex=0.8)

```

Subsampling may lead to loose an important part of 
the original signal.
This loss may vary according to the type of
observed variable 
(for example, several peaks of the time series are
missrepresented by the subsampled signal), and
this effect will become stronger when
considering derivatives of higher degrees (see the
first Section of the present vignette).
It can be useful in such a case to resample the time
series at a higher time resolution.
Resampling can be performed with splines as 
exemplified upper for single time series.
Alternative methods may be preferred depending on 
the complexity of the behavior under study, on the 
degree of subsampling, on the level of noise, etc.
Here, this resampling appears efficient enough to obtain
a good reconstruction of the phase space as illustrated
by the following projections $(x,y)$ and $(x,z)$.

```{r, eval = TRUE, fig.show='hold', fig.align='center', fig.width=4, fig.height=4}
xout <- seq(min(tus), max(tus), by = 0.01)
rsplx <- spline(tus, y = datus[,1], method = "fmm", xout = xout)
rsply <- spline(tus, y = datus[,2], method = "fmm", xout = xout)
rsplz <- spline(tus, y = datus[,3], method = "fmm", xout = xout)
# plot resampled
plot(data[,1], data[,2], type='l', xlab = 'x(t)', ylab = 'y(t)')
lines(rsplx$y, rsply$y, type='l', col='green')
legend(3.5,2,c("original", "resampled"), col=c('black', 'green'), lty=1, cex=0.5)
plot(data[,2], data[,3], type='l', xlab = 'x(t)', ylab = 'z(t)')
lines(rsply$y, rsplz$y, type='l', col='green')
legend(-0.5,12,c("original", "resampled"), col=c('black', 'green'), lty=1, cex=0.5)

```

Discrepancies related to the derivatives
can be easily observed considering the derivative
phase portraits.
The performances of resampling is obviously good for variable $y$:

```{r, eval = TRUE, fig.align='center', fig.align='center', fig.width=4, fig.height=4}
# derivatives
# from original signal
drv1x <- drvSucc(tin, data[,1], nDeriv=2)
drv1y <- drvSucc(tin, data[,2], nDeriv=2)
drv1z <- drvSucc(tin, data[,3], nDeriv=2)
# from resampled signal
drv3x <- drvSucc(rsplx$x, rsplx$y, nDeriv=2)
drv3y <- drvSucc(rsply$x, rsply$y, nDeriv=2)
drv3z <- drvSucc(rsplz$x, rsplz$y, nDeriv=2)
# phase portraits plot for y:
# (y, dy/dt) projection of the original signal (in black) and resampled signal (in green)
plot(drv1y$seriesDeriv[,1], drv1y$seriesDeriv[,2], type='l', xlab = 'y', ylab = 'dy/dt')
lines(drv3y$seriesDeriv[,1], drv3y$seriesDeriv[,2], type='l', col='green')
legend(0,-3,c("original", "resampled"), col=c('black', 'green'), lty=1, cex=0.5)

```

Resampling appears less efficient for variables $x$ and $z$:

```{r, eval = TRUE, fig.show='hold', fig.align='center', fig.width=4, fig.height=4}
# phase portraits plot for x and z:
# (x, dx/dt) projection of the original signal (in black) and resampled signal (in green)
plot(drv1x$seriesDeriv[,1], drv1x$seriesDeriv[,2], type='l', xlab = 'x', ylab = 'dx/dt')
lines(drv3x$seriesDeriv[,1], drv3x$seriesDeriv[,2], type='l', col='green')
legend(-3,-5,c("original", "resampled"), col=c('black', 'green'), lty=1, cex=0.5)

# phase portraits plots
# (z, dz/dt) projection of the original signal (in black) and resampled signal (in green)
plot(drv1z$seriesDeriv[,1], drv1z$seriesDeriv[,2], type='l', xlab = 'z', ylab = 'dz/dt')
lines(drv3z$seriesDeriv[,1], drv3z$seriesDeriv[,2], type='l', col='green')
legend(0,-10,c("original", "resampled"), col=c('black', 'green'), lty=1, cex=0.5)

```


## Conclusion and next step

Although resampling with splines may not permit to reconstruct
the original signal perfectly, it is found that resampling
can enable to obtain models producing a dynamical behavior
very similar to the dynamics of the original dynamical
system.
In practice, it can be an efficient way to deal with subsampling.
This type of resampling could be efficiently used for
applying the global modelling technique both from
single^[S. Mangiarotti, L. Drapeau, & C. Letellier, 2014.
Two chaotic global models for cereal crops cycles observed from
satellite in Northern Morocco, *Chaos*, **24**, 023130.],
and multiple time series^[S. Mangiarotti, 2015. Low dimensional
chaotic models for the plague epidemic in Bombay (1896-1911),
*Chaos, Solitons & Fractals*, **81**(A), 184-196.]$^,$ ^[S. Mangiarotti,
M. Peyre & M. Huc, 2016. A chaotic model for the epidemic
of Ebola virus disease in West Africa (2013-2016).
*Chaos*, **26**, 113112.].

Once the time series have been carefully prepared, the global
modelling technique can be applied. This next step will be
presented in the following vignette `III_Modelling`.
